---
title: "Differences in the effects of exercise on cognition based on sleep quality in people with chronic stroke"
author: "RSF"
date: "2023-06-01"
output: 
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# 1) Data set-up and management

Load Packages
```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(readxl, plyr, lme4,nlme, robustlmm, car, broom, lsmeans, ggplot2, psych, HLMdiag, tableone,mice,mediation)
```

Load Data
```{r}
setwd("~/Desktop/Manuscripts/Ongoing Projects/Vitality PA and Sleep")#Directory where you put the spreadsheet
Vitality <- read_excel("VITALITY_MASTER_FINAL_SLEEP_September26,2023.xlsx")
```

Clean data for analysis
```{r}
spaceless <- function(x) {colnames(x) <- gsub(" ", "_", colnames(x));x}
data1<-spaceless(Vitality)
data2<-data1[c(1:7,9:25,27,29,31,35:37,44,45,50,51,53:56,63,64:72,74,75,76,78:97)]
library(dplyr)
data3<-rename(data2, c("ID"="Participant_ID", "Time"="Timepoint", "Type_of_stroke"="Type_of_stroke_(first_stroke)", "Hemisphere_affected"="Hemisphere_affected_(first_stroke)", 
                       "Location_of_stroke"="Location_of_stroke_(first_stroke)","DSF"="Digit_Span_Forward_Total_Correct", "DSB"="Digit_Span_Backward_Total_Correct",
                       "Trails_A"="Trail_Making_A_time_to_complete", "Trails_B"="Trail_Making_B_time_to_complete", "Trails_B_A"="Trail_Making_Test_Score_(B-A)",
                       "WHR"="Waist_to_Hip_Ratio", "ABC_Score"="Average_ABC_Score", "FCI_Total"="FCI_Item_Total", "Total_IADL"="Total_IADL_Score", "SPPB_Score"="NIA_Total_Score",
                       "RH_Grip_Status"="Unable_to_test_Right_Hand_Strength", "RH_Grip_Strength"="Right_Hand_Mean_Kg", "LH_Grip_Status"="Unable_to_test_Left_Hand_Strength",
                       "LH_Grip_Strength"="Left_Hand_Mean_Kg","RQ_Ext_Status"="Unable_to_test_Right_Quad_Strength","RQ_Ext_Strength"="Mean_Right_Quad_Strength",
                       "LQ_Ext_Status"="Unable_to_test_Left_Quad_Strength","LQ_Ext_Strength"="Mean_Left_Quad_Strength"))
detach("package:dplyr", unload = TRUE)

#Education
data3$EDU<-NULL
data3$EDU[data3$Education==1]<- "High School or Less"
data3$EDU[data3$Education==2]<- "High School or Less"
data3$EDU[data3$Education==3]<- "High School or Less"
data3$EDU[data3$Education==4]<- "Trade School or Some College"
data3$EDU[data3$Education==5]<- "Trade School or Some College"
data3$EDU[data3$Education==6]<- "University Degree or Higher"

#Grip Strength
data3$RH_Grip_Status[data3$RH_Grip_Status=="Unable to Test Right Hand Strength"]<- 0
data3$RH_Grip_Status[is.na(data3$RH_Grip_Status)]<- 1
data3$LH_Grip_Status[data3$LH_Grip_Status=="Unable to Test Left Hand Strength"]<- 0
data3$LH_Grip_Status[is.na(data3$LH_Grip_Status)]<- 1

data3$Avg_Grip_Strength<-NA
data3$Avg_Grip_Strength[data3$RH_Grip_Status==0 & data3$LH_Grip_Status==1]<-data3$LH_Grip_Strength
data3$Avg_Grip_Strength[data3$RH_Grip_Status==1 & data3$LH_Grip_Status==0]<-data3$RH_Grip_Strength
data3$Avg_Grip_Strength[data3$RH_Grip_Status==1 & data3$LH_Grip_Status==1]<-(data3$RH_Grip_Strength + data3$LH_Grip_Strength)/2


#Quad Strength
data3$RQ_Ext_Status[is.na(data3$RQ_Ext_Status)]<- 1
data3$RQ_Ext_Status[data3$RQ_Ext_Status!=1]<-0
data3$LQ_Ext_Status[is.na(data3$LQ_Ext_Status)]<- 1
data3$LQ_Ext_Status[data3$RQ_Ext_Status!=1]<-0

data3$Avg_Quad_Strength<-NA
data3$Avg_Quad_Strength[data3$RQ_Ext_Status==0 & data3$LQ_Ext_Status==1]<-data3$LQ_Ext_Strength
data3$Avg_Quad_Strength[data3$RQ_Ext_Status==1 & data3$LQ_Ext_Status==0]<-data3$RQ_Ext_Strength
data3$Avg_Quad_Strength[data3$RQ_Ext_Status==1 & data3$LQ_Ext_Status==1]<-(data3$LQ_Ext_Strength + data3$RQ_Ext_Strength)/2


#Stroop Interference Scoring
data3$Stroop_INT_Ratio<- (data3$Median.inc.RT - data3$Median.con.RT)/data3$Median.neut.RT
data3$Stroop_INT_Diff<- data3$Median.inc.RT - data3$Median.con.RT
data3$Stroop_IIV<- data1$SD.inc.RT - data1$SD.con.RT

#Digit Span Forward-Backward
data3$DSFB<- data3$DSF - data3$DSB

#Gait Speed
data3$GaitSpeed<- 4/data1$NIA_Walking_Time_in_seconds

#walking Aid Use
data1$Aid<- NA
data3$Aid[data1$Walking_Aid=="None" & data1$Timepoint == 1]<-0
data3$Aid[data1$Walking_Aid=="AFO on left" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="AFO on left, and needs transfer belt" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="arm resst used" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Brace" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Cane" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Cane\r\nBrace" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Cane\r\nBrace (R-leg brace)\r\nBrace" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Cane\r\nBrace\r\nNeurostep implant in leg" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Cane\r\nBrace\r\nUsed armrests" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Cane\r\nHand push off chair" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Cane\r\nHand push off from chair" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Cane\r\nHand push off of chair" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Cane\r\nHands push off chair" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Nordic Poles" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="support belt for back" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Walker" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Walker Cane combination" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="walking pole (only one) - left side" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="walking poles" & data1$Timepoint == 1]<-1
data3$Aid[data1$Walking_Aid=="Walking poles" & data1$Timepoint == 1]<-1

#Remove ID VIT_083
data4<-subset(data3, ID!="VIT_083")

data4$OSA<-as.character(data4$OSA)
```

Determine participants with missing sleep data at baseline for PSQI(N=0), MW Duration (N=11), and MW Efficiency (N=11)
```{r}
#Exclude Participants without Sleep Data at Baseline
library(dplyr)
Sleep_data_1 <- data4 %>% 
  group_by(ID) %>%
  mutate(Missing_PSQI = as.factor(ifelse(Time==1 & !is.na(PSQI),1,0)),
         Missing_Duration = as.factor(ifelse(Time==1 & !is.na(MW_Duration),1,0)),
         Missing_Efficiency = as.factor(ifelse(Time==1 & !is.na(MW_Efficiency),1,0))
         )
baseline_missing<-subset(Sleep_data_1, Time==1)

#Missing PSQI
table(baseline_missing$Missing_PSQI)
#Missing MW Duration
table(baseline_missing$Missing_Duration)
#Missing MW Efficiency
table(baseline_missing$Missing_Efficiency)
```


Group participants into "good sleepers" and "bad sleepers"
```{r}
Sleep_data_2<-data4 %>%
  group_by(ID) %>%
  mutate(Good_Sleep_Quality = as.factor(ifelse(Time==1 & PSQI<=5,1,0)),
         Good_Sleep_Duration = as.factor(ifelse(Time==1 & MW_Duration>=420 & MW_Duration<=490,1,0)),
         Good_Sleep_Efficiency = as.factor(ifelse(Time==1 & MW_Efficiency>=85,1,0))
         )
```

# 3) Baseline Participant Characteristics by group and sleep

```{r}
Baseline<-subset(Sleep_data_2, Time==1)
GoodSleep<-subset(Baseline, Good_Sleep_Quality==1)
BadSleep<-subset(Baseline, Good_Sleep_Quality==0)
GoodSleepDur<-subset(Baseline, Good_Sleep_Duration==1)
BadSleepDur<-subset(Baseline, Good_Sleep_Duration==0)
GoodSleepEff<-subset(Baseline, Good_Sleep_Efficiency==1)
BadSleepEff<-subset(Baseline, Good_Sleep_Efficiency==0)

vars<-dput(names(Baseline[c(31,32,34,72,4,6,7,36,37,23,24,55,54,14,71,39,79,61,56,57)]))
```

## 3.1) All participants
```{r}
Table1_Full<-CreateTableOne(vars=vars, data=Baseline)
print(Table1_Full,contDigits=2,missing=TRUE,quote=TRUE)
```

## 3.2) All participants stratified by group
```{r}
Table1_Full<-CreateTableOne(vars=vars, strata = "Group", data=Baseline)
print(Table1_Full,contDigits=2,missing=TRUE,quote=TRUE)
```

## 3.3) Participants categorized by subjective sleep quality 
(i.e., Good subjective sleep PSQI ≤ 5)
```{r}
Supp_Table1a<-CreateTableOne(vars=vars, strata = "Group", data=GoodSleep)
print(Supp_Table1a,contDigits=2,missing=TRUE,quote=TRUE)

Supp_Table1b<-CreateTableOne(vars=vars, strata = "Group", data=BadSleep)
print(Supp_Table1b,contDigits=2,missing=TRUE,quote=TRUE)
```

## 3.4) Participants categorized by sleep duration 
(i.e., Good sleep duration 420-490 minutes/night)
```{r}
Supp_Table2a<-CreateTableOne(vars=vars, strata = "Group", data=GoodSleepDur)
print(Supp_Table2a,contDigits=2,missing=TRUE,quote=TRUE)

Supp_Table2b<-CreateTableOne(vars=vars, strat = "Group", data=BadSleepDur)
print(Supp_Table2b,contDigits=2,missing=TRUE,quote=TRUE)
```

## 3.5) Participants categorized by sleep efficiency 
(i.e., Good sleep efficiency ≥ 85%/night)
```{r}
Supp_Table3a<-CreateTableOne(vars=vars, strata = "Group", data=GoodSleepEff)
print(Supp_Table3a,contDigits=2,missing=TRUE,quote=TRUE)

Supp_Table3b<-CreateTableOne(vars=vars, strat = "Group", data=BadSleepEff)
print(Supp_Table3b,contDigits=2,missing=TRUE,quote=TRUE)
```

# 4) Primary Linear Mixed Model 

Note: This model is similar to that of the primary model from Liu-Ambrose et al., 2022. The primary difference is that the time points included are baseline, three months and six months, rather than baseline, six months and 12 months. This is done because sleep was onl

Prepare data for linear mixed model
```{r}
data5<-Sleep_data_2[c(1:3,81:83,31,32,34,72,4,6,7,36,37,23,24,55,54,14,71,39,79)]

library(dplyr)
data5 <-rename(data5, ADAS_COG_TOTAL = ADAS_COG_11)

varying<-data5[c(1:2,20:23:length(data5))]
baseline<-subset(data5[-c(20:23:length(data5))], Time==1)

varying.1 <- subset(varying,Time==1)
varying.2 <- subset(varying,Time==3)
varying.3 <- subset(varying,Time==4)

colnames(varying.1) <- paste(colnames(varying.1),"1",sep=".")
colnames(varying.2) <- paste(colnames(varying.2),"2",sep=".")
colnames(varying.3) <- paste(colnames(varying.3),"3",sep=".")

wide.data <- left_join(baseline,varying.1,by=c("ID"="ID.1")) %>% 
  left_join(.,varying.2,by=c("ID"="ID.2")) %>% 
  left_join(.,varying.3,by=c("ID"="ID.3"))
detach("package:dplyr", unload = TRUE)

Timevars <- grep("Time",colnames(wide.data),value=TRUE)
wide.data <- wide.data[,!(colnames(wide.data)%in%Timevars)]

colnames(wide.data) <- (gsub("_","",colnames(wide.data)))
colnames(wide.data) <- (gsub(".1","baseline",colnames(wide.data)))

data7 <- reshape(as.data.frame(wide.data),idvar="ID",varying=23:30,direction="long",sep=".") #reshape to long data frame with T2 and T3 repeated outcome and baseline as separate time-invariant

data7$Sex<- ifelse(data7$Gender=="Male",1,0)

data7$Group.1<-NA
data7$Group.1[data7$Group=="COG"]<-1
data7$Group.1[data7$Group=="BAT"]<-0
data7$Group.1[data7$Group=="EX"]<-2

#LMM function for comparing Exercise versus Control and Cognitive versus Control (USED FOR JAMA PAPER)
lmm.ex.mdl<- function(y,cov1,cov2,cov3){
  model <- lmer(as.formula(paste0(y, "~factor(Group.1)*factor(time) +", cov1, "+", cov2, "+", cov3, "+ (1|ID)")), data= data7)
  modelsummary <- summary(model)
  means.1 <- lsmeans(model, ~Group.1|time)
  contrasts <- contrast(means.1, "trt.vs.ctrl", adj="dunnet")
  conf.95<-confint(contrasts, parm, level = 0.95)
  return(list(summary=modelsummary, lsmeans=means.1, group_contrasts=contrasts, contrasts_ci.95=conf.95))
}

emm_options(opt.digits = FALSE)
```

## 4.1) Primary Analysis: ADAS-Cog Plus
```{r}
lmm.ex.mdl("ADASCOGPLUS", "ADASCOGPLUSbaseline", "MMSE", "FMTotalScore")#PRIMARY OUTCOME:ADAS-COG Plus
```

## 4.2) Secondary Outcome: ADAS-Cog 13
```{r}
lmm.ex.mdl("ADASCOGTOTAL", "ADASCOGTOTALbaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Stroop Interference Ratio
```

## 4.3) Secondary Outcome: SPPB
```{r}
lmm.ex.mdl("SPPBScore", "SPPBScorebaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Short Physical Performance Battery
```

## 4.4) Secondary Outcome: Gait Speed
```{r}
lmm.ex.mdl("GaitSpeed", "GaitSpeedbaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Gaitspeed
```


# 5) Linear mixed model for comparing outcomes based on subjective sleep quality (Good sleep quality PSQI ≤ 5)

Model
```{r}
lmm.ex2.mdl<- function(y,cov1,cov2,cov3){
  model <- lmer(as.formula(paste0(y, "~factor(Group.1)*factor(time)*factor(GoodSleepQuality)","+", cov1, "+", cov2, "+", cov3, "+ (1|ID)")), data= data7)
  modelsummary <- summary(model)
  means.1 <- lsmeans(model, ~Group.1|GoodSleepQuality|time)
  contrasts <- contrast(means.1, "trt.vs.ctrl", adj="dunnet")
  conf.95<-confint(contrasts, parm, level = 0.95)
  return(list(summary=modelsummary, lsmeans=means.1, group_contrasts=contrasts, contrasts_ci.95=conf.95))
}
```

## 5.1) Primary Analysis: ADAS-Cog Plus
```{r}
lmm.ex2.mdl("ADASCOGPLUS","ADASCOGPLUSbaseline","MMSE", "FMTotalScore")#PRIMARY OUTCOME:ADAS-COG Plus
```

## 5.2) Secondary Outcome: ADAS-Cog 13
```{r}
lmm.ex2.mdl("ADASCOGTOTAL", "ADASCOGTOTALbaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Stroop Interference Ratio
```

## 5.3) Secondary Outcome: SPPB
```{r}
lmm.ex2.mdl("SPPBScore", "SPPBScorebaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Short Physical Performance Battery
```

## 5.4) Secondary Outcome: Gait Speed
```{r}
lmm.ex2.mdl("GaitSpeed", "GaitSpeedbaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Gaitspeed
```


# 6) Linear mixed model for comparing outcomes based on objective sleep duration (Good sleep duration = 420-490 minutes/night)

Model
```{r}
lmm.ex3.mdl<- function(y,cov1,cov2,cov3){
  model <- lmer(as.formula(paste0(y, "~factor(Group.1)*factor(time)*factor(GoodSleepDuration)","+", cov1, "+", cov2, "+", cov3, "+ (1|ID)")),  data= data7)
  modelsummary <- summary(model)
  means.1 <- lsmeans(model, ~Group.1|GoodSleepDuration|time)
  contrasts <- contrast(means.1, "trt.vs.ctrl", adj="dunnet")
  conf.95<-confint(contrasts, parm, level = 0.95)
  return(list(summary=modelsummary, lsmeans=means.1, group_contrasts=contrasts, contrasts_ci.95=conf.95))
}
```

## 6.1) Primary Analysis: ADAS-Cog Plus
```{r}
lmm.ex3.mdl("ADASCOGPLUS","ADASCOGPLUSbaseline","MMSE", "FMTotalScore")#PRIMARY OUTCOME:ADAS-COG Plus
```

## 6.2) Secondary Outcome: ADAS-Cog 13
```{r}
lmm.ex3.mdl("ADASCOGTOTAL", "ADASCOGTOTALbaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Stroop Interference Ratio
```

## 6.3) Secondary Outcome: SPPB
```{r}
lmm.ex3.mdl("SPPBScore", "SPPBScorebaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Short Physical Performance Battery
```

## 6.4) Secondary Outcome: Gait Speed
```{r}
lmm.ex3.mdl("GaitSpeed", "GaitSpeedbaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Gaitspeed
```


# 7) Linear mixed model for comparing outcomes based on objective sleep efficiency (Good sleep efficiency ≥ 85%/night)

Model
```{r}
lmm.ex4.mdl<- function(y,cov1,cov2,cov3){
  model <- lmer(as.formula(paste0(y, "~factor(Group.1)*factor(time)*factor(GoodSleepEfficiency)","+", cov1, "+", cov2, "+", cov3, "+ (1|ID)")), data= data7)
  modelsummary <- summary(model)
  means.1 <- lsmeans(model, ~Group.1|GoodSleepEfficiency|time)
  contrasts <- contrast(means.1, "trt.vs.ctrl", adj="dunnet")
  conf.95<-confint(contrasts, parm, level = 0.95)
  return(list(summary=modelsummary, lsmeans=means.1, group_contrasts=contrasts, contrasts_ci.95=conf.95))
}
```

## 7.1) Primary Analysis: ADAS-Cog Plus
```{r}
lmm.ex4.mdl("ADASCOGPLUS","ADASCOGPLUSbaseline","MMSE", "FMTotalScore")#PRIMARY OUTCOME:ADAS-COG Plus
```

## 7.2) Secondary Outcome: ADAS-Cog 13
```{r}
lmm.ex4.mdl("ADASCOGTOTAL", "ADASCOGTOTALbaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Stroop Interference Ratio
```

## 7.3) Secondary Outcome: SPPB
```{r}
lmm.ex4.mdl("SPPBScore", "SPPBScorebaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Short Physical Performance Battery
```

## 7.4) Secondary Outcome: Gait Speed
```{r}
lmm.ex4.mdl("GaitSpeed", "GaitSpeedbaseline", "MMSE", "FMTotalScore")#SECONDARY OUTCOME: Gaitspeed
```